---
title: "Sentiment Analysis of Indian Political Tweets: 2019 General Elections"
author: "Piyush Zaware"
output:
  github_document:
    toc: true
    number_sections: true
---

## INTRODUCTION
During the 2019 Indian General Elections, I noticed how significantly Twitter transformed into a political battleground. Political parties and their supporters increasingly turned to the platform to influence discourse, shape narratives, and connect with voters. In this project, I analyze a large collection of tweets posted between January and May 2019, with the aim of understanding how sentiment and emotional tone varied across different political parties. Using tidy text analysis, I focus on word frequencies, sentiment patterns, and emotional vocabulary to explore the linguistic footprint of Indian political campaigning on social media.

```{r setup, include=FALSE}
library(tidyverse)
library(tidytext)
library(lubridate)
library(stringr)
library(scales)
library(ggthemes)
library(textdata)
library(wordcloud)
library(reshape2)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## DATA DESCRIPTION

The dataset I used consists of tweets collected during the election campaign window—from January 1 to May 23, 2019. These tweets were extracted based on keywords, hashtags, and usernames linked to political parties and leaders. Each entry includes the tweet content, timestamp, and the Twitter handle of the user who posted it. To narrow the scope, I filtered the dataset to focus only on tweets from the official campaign period and removed tweets that were empty or improperly formatted.

```{r data importing and preprocessing}
tweets <- read_csv("IndianElection19TwitterData.csv")
tweets <- tweets %>%
  rename(text = Tweet, timestamp = Date, user = User) %>%
  mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%d %H:%M:%S"),
         date = as.Date(timestamp)) %>%
  filter(date >= as.Date("2019-01-01") & date <= as.Date("2019-05-23")) %>%
  drop_na(text)
```
```{r cleaning}
# Unnest words and remove stopwords
data("stop_words")

tweet_words <- tweets %>%
  select(user, date, text) %>%
  unnest_tokens(word, text) %>%
  filter(!word %in% stop_words$word, 
         str_detect(word, "^[a-z']+$"))
```

## DATA CLEANING AND PREPROCESSING

Before diving into the analysis, I cleaned the data to prepare it for text mining. I tokenized the tweets into individual words, converted them to lowercase, removed punctuation, and eliminated common stop words such as “the” or “and.” This preprocessing step allowed me to focus on politically meaningful words that reveal discourse patterns, rather than on grammatical fillers.

## WORD FREQUENCY ANALYSIS

To begin, I identified the most common words across all tweets. As expected, names like “modi,” “rahul,” “bjp,” “vote,” and “congress” topped the list. These high-frequency words confirm the dataset’s strong political focus and show how central party figures were to the online discussion. It also reflects the dominance of national issues and personalities in voters’ minds during the campaign.

To go beyond basic word frequency, I calculated TF-IDF scores to find unique words used by individual users. This technique helps spotlight the most distinctive words for each handle—whether political parties, media houses, or individuals. For instance, some users used words like “chowkidar,” “priyanka,” or “surgicalstrike” far more often than others, suggesting targeted narratives and ideological leanings. It also gave me a sense of who was pushing which messages more aggressively.

```{r frequently used words}
top_words <- tweet_words %>%
  count(word, sort = TRUE) %>%
  top_n(15)

ggplot(top_words, aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Most Common Words in Political Tweets (2019)",
       x = NULL, y = "Frequency") +
  theme_minimal()
```
```{r vocab}
user_tf_idf <- tweet_words %>%
  count(user, word) %>%
  bind_tf_idf(word, user, n) %>%
  arrange(desc(tf_idf))

top_tf_idf <- user_tf_idf %>%
  group_by(user) %>%
  slice_max(tf_idf, n = 1) %>%
  ungroup()

head(top_tf_idf, 10)
```

## SENTIMENT TRAJECTORY

Using the Bing sentiment lexicon, I calculated daily sentiment scores—by subtracting the number of negative words from positive ones. When I plotted these over time, I noticed spikes in both directions around mid-April and early May, which align with key phases of the election campaign. While there were brief sentiment surges, the overall tone remained relatively balanced across the period. This suggested to me that praise and criticism were flowing in almost equal measure during the campaign, possibly driven by debates, manifesto releases, and campaign rallies.

```{r sentiment}
# Use Bing sentiment lexicon
sentiment_words <- tweet_words %>%
  inner_join(get_sentiments("bing"))

sentiment_by_day <- sentiment_words %>%
  count(date, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(net_sentiment = positive - negative)

ggplot(sentiment_by_day, aes(x = date, y = net_sentiment)) +
  geom_line(color = "darkred") +
  labs(title = "Daily Net Sentiment of Political Tweets",
       x = "Date", y = "Net Sentiment") +
  theme_minimal()
```

## SENTIMENT WORD CLOUD

To visualize sentiment-rich words, I generated a word cloud using terms matched to either positive or negative sentiment. Words like “development,” “trust,” and “victory” stood out on the positive side, while “lies,” “corrupt,” and “scam” featured among the negative. This mix of aspirational and accusatory language felt typical of high-stakes electoral contests, especially in India’s charged political climate.

```{r sentiment word cloud}
wordcloud_data <- sentiment_words %>%
  count(word, sentiment, sort = TRUE)

wordcloud(words = wordcloud_data$word, 
          freq = wordcloud_data$n, 
          min.freq = 30,
          max.words = 100,
          colors = c("firebrick", "forestgreen"),
          random.order = FALSE)
```
```{r second project fun}
nrc <- get_sentiments("nrc")
```

```{r bing sentiment}
bing <- get_sentiments("bing")
```

## PARTYWISE CLASSIFICATION

To better compare patterns, I assigned tweets to major political parties based on keyword mentions (like “Modi” for BJP or “Rahul Gandhi” for INC). This classification enabled me to break down sentiment and emotional tones by party. I filtered out tweets without clear party references to maintain clarity in comparison.

```{r partywise}
# Party keywords
party_patterns <- list(
  BJP   = c("@BJP4India", "\\bBJP\\b", "Modi", "Narendra Modi"),
  INC   = c("@INCIndia", "\\bCongress\\b", "Rahul Gandhi", "@RahulGandhi"),
  AAP   = c("@AamAadmiParty", "\\bAAP\\b", "Arvind Kejriwal"),
  TMC   = c("@AITCofficial", "\\bTMC\\b", "Mamata Banerjee"),
  CPIM  = c("@cpimspeak", "@cpofindia", "\\bCPI\\b", "\\bCPIM\\b"),
  SP    = c("@samajwadiparty", "Akhilesh Yadav", "\\bSP\\b")
)

# Assign party mentions
detect_party <- function(tweet) {
  for (party in names(party_patterns)) {
    if (any(str_detect(tweet, regex(party_patterns[[party]], ignore_case = TRUE)))) {
      return(party)
    }
  }
  return("Other")
}

tweets <- tweets %>%
  mutate(party = map_chr(text, detect_party)) %>%
  filter(party != "Other")  # Optional: focus only on clear party mentions

# Tokenize and clean
tidy_tweets <- tweets %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words, by = "word") %>%
  filter(str_detect(word, "^[a-z]+$"))  # remove numbers/symbols

```

## SENTIMENT BY PARTY

The party-wise sentiment analysis showed interesting contrasts. Tweets about the BJP contained slightly more positive sentiment compared to others—possibly reflecting successful outreach or high engagement from supporters. On the other hand, tweets about the Congress and AAP parties had a more critical tone, with higher levels of negative sentiment. I interpreted this as evidence of an opposition-driven strategy that emphasized critique rather than celebration.

```{r sentiment analysis using bing}
# Sentiment analysis using Bing
sentiment_bing <- tidy_tweets %>%
  inner_join(bing, by = "word") %>%
  count(party, sentiment, sort = TRUE)

ggplot(sentiment_bing, aes(x = reorder(party, -n), y = n, fill = sentiment)) +
  geom_col(position = "dodge") +
  labs(title = "Sentiment Word Usage by Party",
       x = "Party", y = "Word Count", fill = "Sentiment") +
  theme_minimal()
```

## EMOTIONAL BREAKDOWN

Using the NRC lexicon, I categorized emotion-related words (e.g., joy, anger, fear, trust). BJP-related tweets leaned more heavily into emotions like “joy” and “trust,” while INC and TMC were more frequently associated with “anger” and “fear.” This variation may reflect differing campaign tactics—BJP projecting confidence and momentum, while the opposition appealed to dissatisfaction and a sense of crisis.

```{r emotion breakdwon using nrc}
# Emotion breakdown using NRC
emotion_df <- tidy_tweets %>%
  inner_join(nrc, by = "word") %>%
  count(party, sentiment) %>%
  filter(sentiment %in% c("joy", "anger", "fear", "trust", "sadness", "disgust"))

ggplot(emotion_df, aes(x = reorder(party, -n), y = n, fill = sentiment)) +
  geom_col(position = "stack") +
  labs(title = "Emotion Distribution Across Parties (NRC Lexicon)",
       x = "Party", y = "Word Count", fill = "Emotion") +
  theme_minimal()
```

## TOP WORDS BY EACH PARTY

By comparing the top 10 words used in tweets about each party, I could see how their messaging differed. BJP tweets often referenced “nation,” “development,” and “leadership.” Congress-related tweets focused more on terms like “farmer,” “NYAY,” and “jobs,” which echoed their welfare agenda. AAP’s vocabulary highlighted governance priorities, with words like “education” and “health.” These differences felt consistent with how the parties positioned themselves to voters—nationalism and growth vs. welfare and governance.

```{r word by party}
# Top words per party
top_words <- tidy_tweets %>%
  count(party, word, sort = TRUE) %>%
  group_by(party) %>%
  slice_max(order_by = n, n = 10)

ggplot(top_words, aes(x = reorder_within(word, n, party), y = n, fill = party)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~party, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(title = "Top 10 Words Used per Party", x = NULL, y = "Frequency") +
  theme_minimal()
```

## CONCLUSION

Through this text analysis, I gained valuable insights into how different political parties shaped public discourse during the 2019 Indian General Elections. Although all parties used both positive and negative rhetoric, the tone, vocabulary, and emotional triggers varied. BJP appeared to lead in projecting positive sentiment and trust, while Congress and AAP emphasized critique and policy issues. These patterns reflect both top-down campaign messaging and bottom-up voter engagement. If I had more time, I would compare this dataset to offline speech transcripts or newspaper headlines to test whether the online discourse matched real-world narratives.